Four decades of water quality change in the upper San Francisco Estuary

### 2017-01-10 sftrends_manu.Rnw
observed data results done

@@

source('R/funcs.r')
library(WRTDStidal)
library(ggplot2)
library(dplyr)
library(gridExtra)
library(grid)
library(purrr)
library(tidyr)
library(Hmisc)
library(ggrepel)
library(ggtree)
library(forcats)
@

@@

% censored percentages
<<echo = F, cache = T>>=
data(mods)
cvals <- mutate(mods, 
  cper = map(data, function(x){

@@

  })) %>% 
  select(Site_Code, resvar, cper) %>% 
  unnest %>% 
  arrange(-cper) %>% 
  filter(cper > 3)
@

@@

<<echo = F>>=
data(delt_dat)

data(mods)

obs_sm <- select(delt_dat, Site_Code, Location, Date, nh, no23, din) %>% 
  gather('var', 'val', nh:din) %>% 
  group_by(Location, var) %>% 
  summarise(val = mean(val, na.rm = T))

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate whereas ammonium was present at lower concentrations.  However, C3 had a majority of \ac{din} composed of ammonium and others had larger concentrations ammonium relative to nitrite/nitrate depending on the time of year (e.g., P8, D26).  By location, observed concentrations decreased on average from the upper Delta (C3, C10, MD10, P8), to the middle (D19, D26, D28), and to the Suisun stations(D4, D6, D7).

averages by location
averages by years, seasons
seasonality vs no seasonality
ammonium v nitrite/nitrate ratio, overall and by season
spikes

@@

>>>>

% for unit macros
\newcommand{\mgl}{mg L$^{-1}$}
\newcommand{\mugl}{$\mu$g L$^{-1}$}

@@

source('R/funcs.r')
library(WRTDStidal)
library(tidyverse)
library(gridExtra)
library(grid)
library(Hmisc)
library(ggrepel)
library(ggtree)
library(NADA)
@

@@

% censored percentages
<<echo = F, cache = T>>=
data(mods)
cvals_all <- mutate(mods, 
  cper = map(data, function(x){

@@

  })) %>% 
  select(Site_Code, resvar, cper) %>% 
  unnest

cvals <- arrange(cvals_all, -cper) %>% 
  filter(cper > 3)
@

@@

% mle summaries of obs data
<<echo = F, cache = T>>=
data(mods)

# averages across Location
obslo <- select(mods, Location, resvar, data) %>% 
  filter(resvar == 'din') %>% 
  unnest %>% 
  select(Location, resvar, res, not_cens) %>% 
  na.omit %>% 
  group_by(Location) %>%
  mutate(res = exp(res)) %>% 
  nest %>% 
  mutate(
    ests = map(data, function(x){
      cenmle(x$res, !x$not_cens, conf.int = 0.95) %>%   
        mean %>% 
        as.list
    })
  ) %>% 
  select(-data) %>% 
  mutate(ests = map(ests, bind_rows)) %>% 
  unnest %>% 
  split(., .$Location)

# averages by site
obssi <- select(mods, Site_Code, resvar, data) %>% 
  unnest %>% 
  select(Site_Code, resvar, res, not_cens) %>% 
  na.omit %>% 
  group_by(Site_Code, resvar) %>%
  mutate(res = exp(res)) %>% 
  nest %>% 
  mutate(
    ests = map(data, function(x){
      cenmle(x$res, !x$not_cens, conf.int = 0.95) %>% 
        mean %>% 
        as.list
    })
  ) %>% 
  select(-data) %>% 
  mutate(ests = map(ests, bind_rows)) %>% 
  unnest %>% 
  group_by(resvar) %>% 
  filter(mean %in% range(mean)) %>% 
  arrange(resvar, -mean) %>% 
  split(., .$resvar)

# averages across years
obsyr <- select(mods, Site_Code, resvar, data) %>% 
  unnest %>% 
  select(Site_Code, resvar, date, res, not_cens) %>% 
  na.omit %>% 
  mutate(
    res = exp(res), 
    yr = lubridate::year(date), 
    yr = ifelse(yr < 1996, '1976-1995', '1996-2013') 
    ) %>% 
  group_by(resvar, yr) %>% 
  nest %>% 
  mutate(
    ests = map(data, function(x){
      cenmle(x$res, !x$not_cens, conf.int = 0.95) %>% 
        mean %>% 
        as.list
    })
  ) %>% 
  select(-data) %>% 
  mutate(ests = map(ests, bind_rows)) %>% 
  unnest %>% 
  split(., .$resvar)

# averages by season
obsmo <- select(mods, Site_Code, resvar, data) %>% 
  unnest %>% 
  select(Site_Code, resvar, date, res, not_cens) %>% 
  na.omit %>% 
  mutate(
    res = exp(res), 
    mo = lubridate::month(date), 
    mo = factor(mo), 
    mo = fct_collapse(mo, 
      Spring = c('3', '4', '5'),
      Summer = c('6', '7', '8'), 
      Fall = c('9', '10', '11'), 
      Winter = c('12', '1', '2')
      )
    ) %>% 
  group_by(resvar, mo) %>% 
  nest %>% 
  mutate(
    ests = map(data, function(x){
      cenmle(x$res, !x$not_cens, conf.int = 0.95) %>% 
        mean %>% 
        as.list
    })
  ) %>% 
  select(-data) %>% 
  mutate(ests = map(ests, bind_rows)) %>% 
  unnest

@@

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium depending on the time of year.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: \Sexpr{scinot(obslo$Delta['mean'])}$\pm$\Sexpr{scinot(obslo$Delta['se'])} \mgl) and similar for the middle (D19, D26, D28, \Sexpr{scinot(obslo$Middle['mean'])}$\pm$\Sexpr{scinot(obslo$Middle['se'], 3)}) and Suisuan Bay stations (D4, D6, D7, \Sexpr{scinot(obslo$Suisun['mean'])}$\pm$\Sexpr{scinot(obslo$Suisun['se'], 3)}).  Average concentrations were highest at \Sexpr{obssi$din[1, 'Site_Code']} (\Sexpr{scinot(obssi$din[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[1, 'se'])} \mgl) and lowest at \Sexpr{obssi$din[2, 'Site_Code']} (\Sexpr{scinot(obssi$din[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$din[2, 'se'], 3)}) for \ac{din}, highest at \Sexpr{obssi$nh[1, 'Site_Code']} (\Sexpr{scinot(obssi$nh[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[1, 'se'])}) and lowest at \Sexpr{obssi$nh[2, 'Site_Code']} (\Sexpr{scinot(obssi$nh[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$nh[2, 'se'], 3, 3)}) for ammonium, and highest at \Sexpr{obssi$no23[1, 'Site_Code']} (\Sexpr{scinot(obssi$no23[1, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[1, 'se'])}) and lowest at \Sexpr{obssi$no23[2, 'Site_Code']} (\Sexpr{scinot(obssi$no23[2, 'mean'])}$\pm$\Sexpr{scinot(obssi$no23[2, 'se'], 3, 3)}) for nitrite/nitrate. Observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was \Sexpr{scinot(obsyr$din[1, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[1, 'se'])} \mgl for \Sexpr{obsyr$din[1, 'yr']}, compared to \Sexpr{scinot(obsyr$din[2, 'mean'])}$\pm$\Sexpr{scinot(obsyr$din[2, 'se'])} for \Sexpr{obsyr$din[2, 'yr']}. Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.  However, observed seasonality patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).

@@

Formal trend analyses of the observed data suggested that

### 2017-01-10 sftrends_manu.tex
observed data results done

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.

averages by location
averages by years, seasons
seasonality vs no seasonality
ammonium v nitrite/nitrate ratio, overall and by season
spikes

>>>>

% mle summaries of obs data

The observed time series for the ten Delta stations had substantial variation in scale among the nitrogen analytes and differences in apparent seasonal trends (\cref{fig:obsdat}).  In general, long-term (inter-annual) trends were not easily observed from the raw data.  Total \ac{din} for most stations was dominated by nitrite/nitrate, whereas ammonium was a smaller percentage of the total.  However, C3 had a majority of \ac{din} composed of ammonium and other stations (e.g., P8, D16) had higher concentrations of ammonium depending on the time of year.  By location, observed concentrations of \ac{din} for the entire time series were higher on average for the upper Delta stations (C3, C10, MD10, P8; maximum likelihood estimation of mean $\pm$ standard error: $1.16$$\pm$$0.03$ \mgl) and similar for the middle (D19, D26, D28, $0.43$$\pm$$0.01$) and Suisuan Bay stations (D4, D6, D7, $0.44$$\pm$$0.01$).  Average concentrations were highest at P8 ($1.63$$\pm$$0.05$ \mgl) and lowest at C3 ($0.4$$\pm$$0.01$) for \ac{din}, highest at P8 ($0.28$$\pm$$0.02$) and lowest at D28 ($0.05$$\pm$$0.003$) for ammonium, and highest at C10 ($1.4$$\pm$$0.04$) and lowest at C3 ($0.15$$\pm$$0.004$) for nitrite/nitrate. Observed concentrations were also higher later in the time series for all analytes.  For example, average \ac{din} across all stations was $0.62$$\pm$$0.01$ \mgl   for 1976-1995, compared to $0.72$$\pm$$0.01$ for 1996-2013. Seasonal changes across all years also suggested that nitrogen concentrations were lower in the summer and higher in the winter.  However, observed seasonality patterns were inconsistent between sites.  For example, site MD10 had distinct seasonal spikes for elevated \ac{din} in the winter, whereas other stations had less prominent variation between years (D6, D7, \cref{fig:obsdat}).
